Pigeons.jl: Distributed sampling from intractable distributions

### 2024-05-27 paper/examples.tex
Starting to go through paper...

\section{Examples}
In this section we present a set of minimal examples 
that demonstrate how to use Pigeons.jl for sampling. 
We also direct readers to our growing list of examples at InferHub 
(\url{https://julia-tempering.github.io/InferHub/}), which hosts a growing collection of posterior 
distributions with an emphasis on difficult (non-log-concave) problems.

@@

\medskip 
In what follows, we demonstrate how to use Pigeons with a Julia Turing model 
applied to a non-identifiable ``coinflip'' data set, modified from the example
at \url{https://turing.ml/v0.22/docs/using-turing/quick-start}. 
The Bayesian model can be formulated as 
\[
  \label{eq:coinflip}
  p_1, p_2 &\stackrel{iid}{\sim} U(0, 1) \\    
  Y \mid p_1, p_2 &\sim \text{Binomial}(n, p_1 p_2).
\]
The random variable $Y$ is the number of heads observed on $n$ coin flips
where the probability of heads is $p_1 p_2$.  
This model is non-identifiable, meaning that it is not possible to distinguish 
the effects of the two different parameters $p_1$ and $p_2$. As a consequence, 
the target distribution exhibits a complicated structure, as displayed in 
\cref{fig:coinflip_posterior}.
The density of interest corresponding to this model is 
\[
  \pi(p_1, p_2) &= \gamma(p_1, p_2)/Z,
\] 
where 
\[ 
  \gamma(p_1, p_2) &= 
    \binom{n}{y} (p_1 p_2)^y (1-p_1 p_2)^{n-y} I[p_1, p_2 \in [0,1]] 
    \label{eq:coinflip_density} \\
  Z &= \int_0^1 \int_0^1 \gamma(p_1, p_2) \, \dee p_1 \, \dee p_2. \label{eq:coinflip_normalization}
\]
The distribution $\pi$ is also known as the \emph{posterior distribution} in 
Bayesian statistics.

@@

\begin{figure*}[t]
    \centering
    \begin{lstlisting}
          --------------------------------------------------------------------------------------
            #scans    restarts      Λ        time(s)    allc(B)  log(Z₁/Z₀)   min(α)     mean(α)
          ---------- ---------- ---------- ---------- ---------- ---------- ---------- --------- 
                  2          0       1.04      0.383   3.48e+07  -4.24e+03          0      0.885
                  4          0       4.06    0.00287   1.79e+06      -16.3   4.63e-06      0.549
                  8          0       3.49    0.00622   3.55e+06      -12.1      0.215      0.612
                 16          0       2.68     0.0161   7.46e+06      -10.2      0.518      0.703
                 32          0       4.29     0.0353   1.37e+07      -11.8      0.222      0.524
                 64          3       3.17     0.0699   2.86e+07      -11.5      0.529      0.648
                128          8       3.56      0.139   5.53e+07      -11.5      0.523      0.605
                256         12       3.38      0.241    1.1e+08      -11.6      0.526      0.625
                512         37       3.48      0.473   2.22e+08        -12      0.527      0.614
           1.02e+03         77       3.55      0.895   4.46e+08      -11.8      0.571      0.605
          --------------------------------------------------------------------------------------
    \end{lstlisting}
    \caption{Standard output provided by Pigeons.jl. 
    Rows indicate tuning rounds of the PT algorithm with an exponentially increasing number 
    of PT iterations (\texttt{\#scans}). Columns indicate various useful diagnostics, 
    such as the number of allocations per round, time (in seconds), and estimates 
    of the log of the normalization constants. 
    The output is described in greater detail in \cref{sec:PT_diagnostics}
    and has been modified to exclude columns that are not described in the paper.}
    \label{fig:standard_out}
\end{figure*}

\subsubsection{Other targets}
As mentioned previously, it is also possible to specify targets with custom 
probability densities, as well as Stan and Turing models. 
Additionally, suppose we have some code implementing vanilla MCMC, written 
in an arbitrary ``foreign'' language such as C++, Python, R, Java, etc. 
Surprisingly, it is very simple to bridge such code with Pigeons.jl. 
The only requirement on the foreign language is that it supports 
reading the standard input and writing to the standard output, 
as described in our online documentation.

@@

\subsubsection{Standard output}
An example of the standard output provided by Pigeons.jl is displayed in \cref{fig:standard_out}. 
Each row of the table in the output indicates a new tuning round in parallel tempering, 
with the \texttt{\#scans} column indicating the number of scans/samples in 
that tuning round. 
During these tuning rounds, Pigeons.jl searches for optimal values of certain PT 
tuning parameters. Other outputs include: 
\begin{itemize}
    \item \texttt{restarts}: a higher number is better. 
    Informally, PT performs well at sampling from high-dimensional and/or multi-modal 
    distributions by pushing samples from an easy-to-sample distribution (the reference) 
    to the more difficult target distribution. 
    A tempered restart happens when a sample from the reference successfully 
    percolates to the target. (See the subsequent sections for a more detailed description 
    of parallel tempering.) 
    When the reference supports i.i.d.~sampling, tempered restarts 
    can enable large jumps in the state space.

@@

\medskip 
To obtain pair plots, we add the 
PairPlots.jl\footnote{\url{https://github.com/sefffal/PairPlots.jl}}
and CairoMakie.jl\footnote{\url{https://github.com/JuliaPlots/CairoMakie.jl}}
packages:
\begin{lstlisting}[language=Julia]
Pkg.add("PairPlots", "CairoMakie")
\end{lstlisting}
and then run
\begin{lstlisting}[language=Julia]
using PairPlots, CairoMakie
my_plot = PairPlots.pairplot(samples) 
display(my_plot)
\end{lstlisting}
The output of the above code chunk with an increased number of samples 
is displayed in \cref{fig:coinflip_posterior}.
\footnote{The code chunk above only works with Julia 1.9.}

>>>>

\section{Examples}
In this section we present a set of minimal examples 
that demonstrate how to use Pigeons.jl for sampling. 
We also direct readers to our growing list of examples at InferHub 
(\url{https://julia-tempering.github.io/InferHub/}), which hosts a  collection of posterior 
distributions with an emphasis on difficult (non-log-concave) problems.

@@

\medskip 
In what follows, we demonstrate how to use Pigeons with a Julia Turing model 
applied to a non-identifiable ``coinflip'' data set.
The Bayesian model can be formulated as 
\[
  \label{eq:coinflip}
  p_1, p_2 &\stackrel{\text{i.i.d.}}{\sim} U(0, 1), \\    
  y \mid p_1, p_2 &\sim \text{Binomial}(n, p_1 p_2).
\]
The random variable $y$ is the number of heads observed on $n$ coin flips
where the probability of heads is $p_1 p_2$.  
This model is non-identifiable, meaning that it is not possible to distinguish 
the effects of the two different parameters $p_1$ and $p_2$. As a consequence, 
the target distribution exhibits a complicated structure, as displayed in 
\cref{fig:coinflip_posterior}.
The density of interest corresponding to this model is 
\[
  \pi(p_1, p_2) &= \gamma(p_1, p_2)/Z,
\] 
where 
\[ 
  \gamma(p_1, p_2) &= 
    \binom{n}{y} (p_1 p_2)^y (1-p_1 p_2)^{n-y} I[p_1, p_2 \in [0,1]] 
    \label{eq:coinflip_density} \\
  Z &= \int_0^1 \int_0^1 \gamma(p_1, p_2) \, \dee p_1 \, \dee p_2. \label{eq:coinflip_normalization}
\]
The distribution $\pi$ is also known as the \emph{posterior distribution} in 
Bayesian statistics.

@@

\begin{figure*}[t]
    \centering
    \begin{lstlisting}
            #scans    restarts        Λ      time(s)    allc(B)  log(Z₁/Z₀)   min(α)     mean(α)

                  2          0       1.04      0.383   3.48e+07  -4.24e+03          0      0.885
                  4          0       4.06    0.00287   1.79e+06      -16.3   4.63e-06      0.549
                  8          0       3.49    0.00622   3.55e+06      -12.1      0.215      0.612
                 16          0       2.68     0.0161   7.46e+06      -10.2      0.518      0.703
                 32          0       4.29     0.0353   1.37e+07      -11.8      0.222      0.524
                 64          3       3.17     0.0699   2.86e+07      -11.5      0.529      0.648
                128          8       3.56      0.139   5.53e+07      -11.5      0.523      0.605
                256         12       3.38      0.241    1.1e+08      -11.6      0.526      0.625
                512         37       3.48      0.473   2.22e+08        -12      0.527      0.614
           1.02e+03         77       3.55      0.895   4.46e+08      -11.8      0.571      0.605

    \end{lstlisting}
    \caption{Standard output provided by Pigeons.jl. 
    Rows indicate tuning rounds of the PT algorithm with an exponentially increasing number 
    of PT iterations (\texttt{\#scans}). Columns indicate various useful diagnostics, 
    such as the amount of memory allocations per round (in bytes), time (in seconds), and estimates 
    of the log of the normalization constants. 
    The output is described in greater detail in \cref{sec:PT_diagnostics}
    and has been modified to exclude columns that are not described in the paper.}
    \label{fig:standard_out}
\end{figure*}

\subsubsection{Other targets}
As mentioned previously, it is also possible to specify targets with custom 
probability densities, as well as Stan and Turing models. 
Additionally, suppose we have some code implementing vanilla MCMC, written 
in an arbitrary ``foreign'' language such as C++, Python, R, Java, etc. 
Surprisingly, it is very simple to bridge such code with Pigeons.jl. 
See \url{https://pigeons.run/stable/input-overview/}.

@@

\subsubsection{Standard output}
An example of the standard output provided by Pigeons.jl is displayed in \cref{fig:standard_out}. 
Each row of the table in the output indicates a new tuning round in parallel tempering, 
with the \texttt{\#scans} column indicating the number of scans/samples in 
that tuning round. 
During these tuning rounds, Pigeons.jl searches for optimal values of certain PT 
tuning parameters. Other outputs include: 
\begin{itemize}
    \item \texttt{restarts}: a higher number is better. 
    Informally, PT propagates samples from an easy-to-sample distribution (the reference) 
    to the more difficult target distribution. 
    A tempered restart happens when a sample from the reference successfully 
    percolates to the target. (See the subsequent sections for a more detailed description 
    of parallel tempering.) 
    When the reference supports i.i.d.~sampling, tempered restarts 
    can enable large jumps in the state space.

@@

\medskip 
To obtain pair plots, we add the 
PairPlots.jl\footnote{\url{https://github.com/sefffal/PairPlots.jl}}
and CairoMakie.jl\footnote{\url{https://github.com/JuliaPlots/CairoMakie.jl}}
packages:
\begin{lstlisting}[language=Julia]
Pkg.add("PairPlots", "CairoMakie")
\end{lstlisting}
and then run
\begin{lstlisting}[language=Julia]
using PairPlots, CairoMakie
my_plot = PairPlots.pairplot(samples) 
display(my_plot)
\end{lstlisting}
The output of the above code chunk with an increased number of samples 
is displayed in \cref{fig:coinflip_posterior}.
\footnote{The code chunk above requires Julia 1.9.}

### 2024-05-27 paper/introduction.tex
Starting to go through paper...

\section{Introduction}
In many scientific application domains, the ability to obtain samples from a 
distribution is crucial. 
For instance, sampling methods have been used to discover magnetic polarization 
in the black hole of galaxy M87 \cite{akiyama2021seven}
and to image the Sagittarius A* black hole \cite{akiyama2022first}.
They have also been used to 
model the evolution of single-cell cancer genomes \cite{salehi2021clonal}, 
infer plasma dynamics inside nuclear fusion reactors \cite{gota2021overview}, 
and to identify gerrymandering in Georgia's 2021 congressional districting plan 
\cite{zhao2022mathematically}.
Similarly, evaluating high-dimensional integrals or sums over complicated 
combinatorial spaces are related tasks that can also be solved with sampling 
via Markov chain Monte Carlo (MCMC) methods. 
However, such calculations can often be bottlenecks in the scientific process, with 
simulations that can last days or even weeks to finish.

\medskip
Pigeons.jl\footnote{The source code for Pigeons.jl can be found at 
\url{https://github.com/Julia-Tempering/Pigeons.jl}. 
Pigeons.jl v0.2.0, used in the example scripts below, is currently compatible with Julia 1.8+.}
enables users to sample efficiently from high-dimensional and complex distributions 
and solve integration problems by 
implementing state-of-the-art sampling algorithms \cite{syed2021nrpt,surjanovic2022vpt} 
that leverage distributed computation. Its simple API allows users to perform such computations 
single-threaded, multi-threaded, and/or distributed over MPI-communicating machines. 
We have stress-tested Pigeons successfully with up to 1000 machines running 
concurrently on a compute cluster (Sockeye) at the University of British Columbia.
Further, Pigeons comes with guarantees on strong parallelism invariance wherein 
the output for a given seed is \emph{identical} irrespective 
of the number of threads or processes. 
Such a level of reproducibility is rare in distributed software but of 
great use for the purposes of debugging in the context of sampling algorithms, 
which produce stochastic output.
Specifically, Pigeons.jl is designed to be suitable and yield reproducible output for:
\begin{enumerate}
    \item one machine running on one thread;
    \item one machine running on several threads;
    \item several machines running, each using one thread, and
    \item several machines running, each using several threads.
\end{enumerate}

@@

    \item Approximating the value of the normalization constant $Z$. For 
    example, in Bayesian statistics, this corresponds to the 
    marginal likelihood, which can be used for model selection. 
\end{enumerate}
Its implementation shines compared to traditional sampling packages in the 
following scenarios:
\begin{itemize}
    \item When the target density $\pi$ is challenging due to a complex structure 
    (e.g., high-dimensional, multi-modal, etc.).

@@

\subsection{What is of interest to the general Julia developer?}
Ensuring code correctness at the intersection of randomized, parallel, and 
distributed algorithms is a challenge. To address this challenge, we designed Pigeons.jl 
based on a principle that we refer to as \textit{strong parallelism invariance}
(SPI).
Namely, the output of Pigeons.jl is completely invariant to the number of machines 
and/or threads.
Without explicitly keeping SPI in mind during software construction, 
the (random) output of the algorithm is only guaranteed to have the same distribution.
This is a much weaker guarantee that, in particular, makes debugging difficult.
However, with our notion of SPI we make debugging and 
software validation considerably easier. This is because the developer can 
first focus on the fully serial randomized algorithm, and then use it as an 
easy-to-compare gold-standard reference for parallel/distributed implementations. 
This strategy is used extensively in Pigeons.jl to ensure correctness. In contrast, 
testing equality in distribution, while possible (e.g., see \cite{geweke2004getting}), 
incurs additional false negatives due to statistical error.

>>>>

\section{Introduction}
In many scientific application domains, the ability to obtain samples from a 
probability distribution is of central interest. 
For instance, sampling methods have been used to discover magnetic polarization 
in the black hole of galaxy M87 \cite{akiyama2021seven}
and to image the Sagittarius A* black hole \cite{akiyama2022first}.
They have also been used to 
model the evolution of single-cell cancer genomes \cite{salehi2021clonal}, 
infer plasma dynamics inside nuclear fusion reactors \cite{gota2021overview}, 
and to identify gerrymandering in Georgia's 2021 congressional districting plan 
\cite{zhao2022mathematically}.
Similarly, evaluating high-dimensional integrals or sums over complicated 
combinatorial spaces are related tasks that can also be solved with sampling 
via Markov chain Monte Carlo (MCMC) methods. 
However, such calculations can often be bottlenecks in the scientific process, with 
simulations that can last days or even weeks to finish.

\medskip
Pigeons.jl\footnote{The source code for Pigeons.jl can be found at 
\url{https://github.com/Julia-Tempering/Pigeons.jl}. 
Pigeons.jl v0.2.0, used in the example scripts below, is currently compatible with Julia 1.8+.}
enables users to sample efficiently from challenging probability distributions 
and solve integration problems by 
implementing state-of-the-art sampling algorithms \cite{syed2021nrpt,surjanovic2022vpt} 
that leverage distributed computation. Its simple user interface facilitates performing such computations 
single-threaded, multi-threaded, and/or distributed over MPI-communicating machines. 
We have stress-tested Pigeons successfully with up to 1000 machines running 
concurrently on a compute cluster (Sockeye) at the University of British Columbia.
Further, Pigeons comes with guarantees on strong parallelism invariance wherein 
the output for a given seed is \emph{identical} irrespective 
of the number of threads or processes. 
Such a level of reproducibility is rare in distributed software but of 
great use for the purposes of debugging in the context of sampling algorithms, 
which produce stochastic output.
Specifically, Pigeons.jl is designed to be suitable and yield reproducible output for:
\begin{enumerate}
    \item one machine running on one thread;
    \item one machine running on several threads;
    \item several machines running, each using one thread, and
    \item several machines running, each using several threads.
\end{enumerate}

\subsection{Latest documentation}

More in-depth documentation and additional examples can be found at: \\ \url{https://pigeons.run/}.

@@

    \item Approximating the value of the normalization constant $Z$. For 
    example, in Bayesian statistics, this corresponds to the 
    marginal likelihood, which can be used for model selection. 
\end{enumerate}
The Pigeons package particularly shines compared to traditional sampling approaches in the 
following scenarios:
\begin{itemize}
    \item When the target density $\pi$ is challenging due to a complex structure 
    (e.g., high-dimensional, multi-modal, etc.).

@@

\subsection{What is of interest to the general Julia developer?}
Ensuring code correctness at the intersection of randomized, parallel, and 
distributed algorithms is a challenge. To address this challenge, we designed Pigeons.jl 
based on a principle that we refer to as \textit{strong parallelism invariance}
(SPI).
Namely, the output of Pigeons.jl is  invariant to the number of machines 
and/or threads.
Without explicitly keeping SPI in mind during software construction, 
the (random) output of the algorithm is only guaranteed to have the same distribution.
This is a much weaker guarantee that, in particular, makes debugging difficult.
However, with our notion of SPI we make debugging and 
software validation considerably easier. This is because the developer can 
first focus on the fully serial randomized algorithm, and then use it as an 
easy-to-compare gold-standard reference for parallel/distributed implementations. 
This strategy is used extensively in Pigeons.jl to ensure correctness. In contrast, 
testing equality in distribution, while possible (e.g., see \cite{geweke2004getting}), 
incurs additional false negatives due to statistical error.

### 2024-05-27 paper/paper.tex
use paper folder
fix
Starting to go through paper...

\usepackage[autonum, blackhypersetup]{shortex}
\usepackage{xcolor}
%\let\subcaption\undefined % overwrite 'subcaption' from the style file
\usepackage[font=scriptsize]{subcaption}
\usepackage{forest}

@@

\pdfsuppresswarningpagegroup=1

@@

We introduce a software package, Pigeons.jl, that provides a way to leverage distributed computation to 
obtain samples from complicated probability distributions, such as 
multimodal posteriors arising in Bayesian inference and high-dimensional 
distributions in statistical mechanics. Pigeons.jl provides simple APIs to 
perform such computations single-threaded, multi-threaded,
and/or distributed over thousands of MPI-communicating machines.
In addition, Pigeons.jl guarantees a property that we call strong parallelism invariance: the 
output for a given seed is identical irrespective of the number of threads and processes, 
which is crucial for scientific reproducibility and software validation. 
We describe the key features of Pigeons.jl and the approach taken to implement 
a distributed and randomized algorithm that satisfies strong parallelism invariance.

@@

\input{src/introduction.tex}
\input{src/examples.tex}
\input{src/PT.tex}
\input{src/SPI.tex}
\input{src/related.tex}
\input{src/conclusion.tex}

>>>>

\usepackage{expl3} % add missing functionality in the old texlive version running on the server 
\usepackage[autonum, blackhypersetup]{shortex}
\usepackage{xcolor}
%\let\subcaption\undefined % overwrite 'subcaption' from the style file
%\usepackage[font=scriptsize]{subcaption}
\usepackage{forest}

@@

\pdfsuppresswarningpagegroup=1 % avoid warning due to pdfs made with photoshop

@@

We introduce a software package, Pigeons.jl, that provides a way to leverage distributed computation to 
obtain samples from complicated probability distributions, such as 
multimodal posteriors arising in Bayesian inference and high-dimensional 
distributions in statistical mechanics. Pigeons.jl provides a simple user interface to 
perform such computations single-threaded, multi-threaded,
and/or distributed over thousands of MPI-communicating machines.
In addition, Pigeons.jl guarantees a property that we call strong parallelism invariance: the 
output for a given seed is identical irrespective of the number of threads and processes, 
which is crucial for scientific reproducibility and software validation. 
We describe the key features of Pigeons.jl and the approach taken to implement 
a distributed and randomized algorithm that satisfies strong parallelism invariance.

@@

\input{introduction.tex}
\input{examples.tex}
\input{PT.tex}
\input{SPI.tex}
\input{related.tex}
\input{conclusion.tex}
